---
title: >
  Configuring `r Biocpkg("iSEE")` apps
author:
- name: Federico Marini
  affiliation: 
  - &id1 Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Center for Thrombosis and Hemostasis (CTH), Mainz
  email: marinif@uni-mainz.de
- name: Aaron Lun
  affiliation: 
  - &id2 Cancer Research UK Cambridge Institute, University of Cambridge
  email: aaron.lun@cruk.cam.ac.uk
- name: Charlotte Soneson
  affiliation: 
  - &id3 Institute of Molecular Life Sciences, University of Zurich
  - SIB Swiss Institute of Bioinformatics
  email: charlottesoneson@gmail.com
- name: Kevin Rue-Albrecht
  affiliation: 
  - &id4 Kennedy Institute of Rheumatology, University of Oxford,
    Headington, Oxford OX3 7FY, UK.
  email: kevin.rue-albrecht@kennedy.ox.ac.uk
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('iSEE')`"
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{3. Configuring iSEE apps}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{iSEE}
  %\VignetteKeywords{GeneExpression, RNASeq, Sequencing, Visualization, QualityControl, GUI}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: iSEE.bib
---

**Compiled date**: `r Sys.Date()`

**Last edited**: 2018-03-08

**License**: `r packageDescription("iSEE")[["License"]]`

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  warning = FALSE,
  message = FALSE
)
stopifnot(requireNamespace("htmltools"))
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
sce <- readRDS("sce.rds")    
```

To load and attach the package to your current workspace, enter:

```{r library}
library(iSEE)
```

# Changing the default start configuration

The default start configuration with one plot of each type may not always be the most appropriate.
`iSEE` allows the user to programmatically modify the initial settings [@kra2018iSEE], avoiding the need to click through the choices to obtain the desired panel setup.
Almost every aspect of the initial app state can be customized, down to whether or not the parameter boxes are open or closed!

To demonstrate this feature, let's assume that we are only interested in feature assay plots.
The default set of panels can be changed via the `initialPanels` argument of the `iSEE` function call.
Given a `SingleCellExperiment` or `SummarizedExperiment` object named `sce`, the following code opens an app with two adjacent feature assay plots.
Note that each panel is set to occupy half the width of the application window, which is set to 12 units by the `r CRANpkg("shiny")` package.

```{r init}
init <- DataFrame(
    Name = c("Feature assay plot 1", "Feature assay plot 2"),
    Width = c(6, 6)
)
app <- iSEE(sce, initialPanels = init)
```

The genes to show on the y-axis in the two plots can be specified via the `featAssayArgs` argument to `iSEE`. This is a `DataFrame`, specifying the initial parameters for each plot.
In this case, we want to modify the `YAxis` and `YAxisFeatName` defaults for the two plots:

```{r fexArg}
fexArg <- featAssayPlotDefaults(sce, 2)
fexArg$YAxisFeatName <- c("0610009L18Rik", "0610009B22Rik")
app <- iSEE(sce, initialPanels = init, featAssayArgs = fexArg)
```

This will open the app with two feature assay plots, showing the selected genes.
Of course, from this starting point, all the interactive functionality is still available, and new panels can be added, modified and linked via point selection.

# Further reading

Many more options can be set to initialise `iSEE` applications in virtually any state and preferred layout.
Users should refer to the following help pages for the full list of values that can be specified in `iSEE`:

- `?redDimPlotDefaults`
- `?colDataPlotDefaults`
- `?featAssayPlotDefaults`
- `?rowDataPlotDefaults,`
- `?rowStatTableDefaults`
- `?sampAssayPlotDefaults`
- `?heatMapPlotDefaults`
- `?"iSEE point parameters"`
- `?"iSEE selection parameters"`

# Session Info {.unnumbered}

```{r sessioninfo}
sessionInfo()
# devtools::session_info()
```

# References {.unnumbered}

