---
title: Deploying custom panels in the `iSEE` interface
author:
- name: Federico Marini
  affiliation: 
  - &id1 Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Center for Thrombosis and Hemostasis (CTH), Mainz
  email: marinif@uni-mainz.de
- name: Aaron Lun
  affiliation: 
  - &id2 Cancer Research UK Cambridge Institute, University of Cambridge
  email: aaron.lun@cruk.cam.ac.uk
- name: Charlotte Soneson
  affiliation: 
  - &id3 Institute of Molecular Life Sciences, University of Zurich
  - SIB Swiss Institute of Bioinformatics
  email: charlottesoneson@gmail.com
- name: Kevin Rue-Albrecht
  affiliation: 
  - &id4 Kennedy Institute of Rheumatology, University of Oxford,
    Headington, Oxford OX3 7FY, UK.
  email: kevin.rue-albrecht@kennedy.ox.ac.uk
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('iSEE')`"
output: 
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{3. Deploying custom panels}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{iSEE}
  %\VignetteKeywords{GeneExpression, RNASeq, Sequencing, Visualization, QualityControl, GUI}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: iSEE.bib
---

**Compiled date**: `r Sys.Date()`

**Last edited**: 2018-03-08

**License**: `r packageDescription("iSEE")[["License"]]`

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    error = FALSE,
    warning = FALSE,
    message = FALSE
)
stopifnot(requireNamespace("htmltools"))
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
sce <- readRDS('sce.rds')
```

# Background

Users can define their own plots or tables to include in the `iSEE` interface [@kra2018iSEE].
These custom panels are intended to receive a subset or rows or columns from other transmitting panels in the interface.
The values in the custom panels will be recomputed "on the fly" by user-defined functions with the transmitted subset.
This provides a flexible and convenient approach for light-weight interactive analysis during data exploration.
For example, selection of a particular subset of samples can be transmitted to a custom plot panel that performs dimensionality reduction on that subset.
Alternatively, the subset can be transmitted to a custom table that performs a differential expression analysis between that subset and all other samples.

# Defining custom functions

## Minimum requirements

Recalculations in custom panels are performed using user-defined functions that are supplied to the `iSEE()` call.
The only requirements are that the function has to accept:

- A `SummarizedExperiment` object or its derivatives as the first argument.
- A vector of row names as the second argument, specifying the rows that are selected in the transmitting _row-panel_ panel.
- A vector of column names as the third argument, specifying the columns that are selected in the transmitting _column-based_ panel.

For custom plot panels, the function is expected to produce a `ggplot` object.
For custom table panels, the function is expected to produce a `data.frame`.

To demonstrate, we'll define an example function `CUSTOM_PCA` below.
This takes a subset of features and cells in a `SingleCellExperiment` object, performs a principal components analysis (PCA) using that subset with `runPCA`, 
and creates a plot of the first two principal components with `plotPCA` [@mccarthy2017scater].

```{r}
library(scater)
CUSTOM_PCA <- function(se, rows, columns, colour_by=NULL, scale_features=TRUE) {
    if (!is.null(columns)) {
        kept <- se[,columns]
    } else {
        return(ggplot())
    }

    scale_features <- as.logical(scale_features)
    kept <- runPCA(kept, feature_set=rows, scale_features=scale_features)
	plotPCA(kept, colour_by=colour_by)
}
```

Note that `rows` and `columns` may be `NULL` if no selection was made in the respective transmitting panels.
How these should be treated is up to the user-defined function.
In the `CUSTOM_PCA` example above, an empty _ggplot_ is returned if there is no selection on the columns, while the default `runPCA` behaviour is used if `feature_set=NULL`.

## Additional arguments

Users of custom panels can specify additional arguments via a text box that accepts multi-line inputs.
This is automatically converted into named arguments before being passed to the custom function.
Each line corresponds to one argument:value pair, and the name of the argument is defined as the string before the first space.

For example, if a user were to type:

```
colour_by Nanog
scale_features FALSE
```

... this would pass `colour_by="Nanog"` and `scale_features="TRUE"` to `CUSTOM_PCA`.
Note that all additional arguments are passed as character strings for the sake of security.
It is the responsibility of the custom function to perform any necessary type checks and conversions (hence the `as.logical` in `CUSTOM_PCA`).

## Example use

Using the `sce` object that we generated in the first vignette, we will add some fields for examining the mean-variance relationship across features:

```{r}
rowData(sce)$mean_log <- rowMeans(logcounts(sce))
rowData(sce)$var_log <- apply(logcounts(sce), 1, var)
```                                                                            

We will then set up an `iSEE` instance with three panels - a reduced dimension plot, a row data plot and a custom plot.
Note how the custom panel initially receives a column selection from the reduced dimension plot and a row selection from the row data plot.
We also set the initial function to the name of `CUSTOM_PCA` in `customDataFun`, and the initial arguments to the values described above.

```{r}
library(iSEE)
reddim <- redDimPlotDefaults(sce, 1)
rowdat <- rowDataPlotDefaults(sce, 1)
rowdat$XAxis <- "Row data"
rowdat$XAxisRowData <- "mean_log"
rowdat$YAxis <- "var_log"

cdp <- customDataPlotDefaults(sce, 1)
cdp$Function <- "CUSTOM_PCA"
cdp$Arguments <- "colour_by Nanog\nscale_features FALSE"
cdp$ColumnSource <- "Reduced dimension plot 1"
cdp$RowSource <- "Row data plot 1"

app <- iSEE(sce, redDimArgs=reddim, rowDataArgs=rowdat, customDataArgs=cdp,
    initialPanels=DataFrame(Name=c("Reduced dimension plot 1", 
            "Row data plot 1", "Custom data plot 1")),
    customDataFun=list(CUSTOM_PCA=CUSTOM_PCA))
```

# Caching for greater efficiency

True R wizards can take advantage of the pass-by-reference activity of environments to cache results throughout the lifetime of the app.
This can be used to avoid repeating time-consuming steps that are not affected by changes in certain parameters.
For example, we could cache the output of `runPCA` to avoid repeating the PCA if only `colour_by` changes.

We will demonstrate this below using a function that computes log-fold changes for the selected features in one subset of samples compared to the others.
For any given column selection, we compute log-fold changes for all features and cache the results in the `caching` environment.
This allows us to avoid recomputing these values if only the row selection changes.

```{r}
caching <- new.env()
CUSTOM_LFC <- function(se, rows, columns) {
    if (is.null(columns)) {
        return(data.frame(logFC=numeric(0)))
    }

    if (!identical(caching$columns, columns)) {
        caching$columns <- columns
        in.subset <- rowMeans(logcounts(sce)[,columns])
        out.subset <- rowMeans(logcounts(sce)[,setdiff(colnames(sce), columns)])
        caching$logFC <- setNames(in.subset - out.subset, rownames(sce))
    } 
       
    lfc <- caching$logFC
    if (!is.null(rows)) {
        out <- data.frame(logFC=lfc[rows], row.names=rows)
    } else {
        out <- data.frame(logFC=lfc, row.names=rownames(se))
    }
    out
}
```

We will then set up an `iSEE` instance with three panels - a reduced dimension plot, a row data plot and a custom statistics table.
This is much the same as described for the custom plot panel.

```{r}
cst <- customStatTableDefaults(sce, 1)
cst$Function <- "CUSTOM_LFC"
cst$ColumnSource <- "Reduced dimension plot 1"
cst$RowSource <- "Row data plot 1"

app2 <- iSEE(sce, redDimArgs=reddim, rowDataArgs=rowdat, customStatArgs=cst, 
    initialPanels=DataFrame(Name=c("Reduced dimension plot 1", 
            "Row data plot 1", "Custom statistics table 1")),
    customStatFun=list(CUSTOM_LFC=CUSTOM_LFC))
```

# Session Info {.unnumbered}

```{r sessioninfo}
sessionInfo()
# devtools::session_info()
```

# References {.unnumbered}


